% $Log: preface.4e.tex,v $
% Revision 1.4  2021/05/17 10:00:00  neil
% *** empty log message ***
%
\parindent=5mm
\chapter{Preface to the 4th Edition}
This 4th edition of ``Programming Algol 68 Made Easy'' or ``PAME''
to its friends has been created to reflect changes in technology and
the a68toc package since Dr. Mountbatten's last version in 2008.

The book, while mostly general in nature, features the Algol-68RS compiler
as a target, through the a68toc Algol 68 to C translator. Examples and
exercises are presented in the dialect supported by a68toc.

To preserve Sian's work to the greatest extent, only very minor
changes have been made to the text: typos, references to older versions
of a68toc and support for more architectures.

The \LaTeX\ source has been updated to use the HyperTex package to
provide internal links suitable for producing a PDF with navigation.
Exercises now have working links to answers and vice versa.

Starting from a version of Dr. Mountbatten's port of a68toc to 32-bit Linux
(v1.15\footnote{While a v1.19 is known, no source of this version has been located}),
the current author has made some advances and released further
versions\footnote{Releases and work-in-progress available on GitHub: \newline\url{https://github.com/coolbikerdad}}.
These include:
\begin{itemize}
\item A complete port to 64-bit systems, removing assumptions (explicit and implied)
about word size in the libraries and the compiler itself.
\item The integration of a well-supported garbage collector (Boehm-Demers-Weiser) as
a compile-time option for the compiler (default for 64-bit systems).
\item Added support for 32-/64-bit ARM processor architectures, removing implicit assumptions about endianness.
\item Added support for 64-bit macOS operating system versions.
\item A fix for a long-standing bug in the compiling of the \verb|BY| part of \verb|FOR| loops.
\item Production of Debian Linux binary packages for i386, amd64, armhf, arm64 architectures.
\end{itemize}

This edition of the book applies to versions of a68toc v1.22 and later.

\bigskip\bigskip
\begin{flushright}
Neil Matthew\\
neil@tilde.co.uk\\
May 2021
\end{flushright}
