\input webmac
% This program is copyright (c) BP65 Phoenix Engineering
\def\Algol{\textbf{Algol 68}}
\def\Web{\textbf{Web68}}

@Introduction.

@ banner = "$Id: params.w68,v 1.4 2003/12/20 11:47:07 sian Exp $"

@@Compiler prelude@
@Modes in the outer reach@
@!NIL!s in the outer reach@
@Names in the outer reach@
@Operators in the outer reach@
@Procedures in the outer reach@
SKIP
@Compiler postlude@


\M1. The local compiler requires a special prelude.

\Y\P$\4\X1:\X\S$\6
\4\\{DECS}\\{params}\\{CONTEXT}\\{VOID}\\{USE}\\{standard}: \37\par
\fi

\M2. And a special postlude. All the public tags go here.

\Y\P$\4\X1:\X\mathrel{+}\S$\6
\\{KEEP}\X15:Public tags\X\\{FINISH}\\{Tokens}.\\{The}\\{basic}\\{mode} \6
\&{for} $\\{tokens}\\{is}\|a\\{structure}\\{which}\\{contains}\\{the}%
\\{characters}$ \1\&{of} $\\{the}\\{token}\W\\{the}!\\{ty}!\\{field}\\{which}%
\\{determines}\\{the}$ \6
\4\&{type} \&{of} $\\{the}\\{token}.\X1:\X=\\{MODE}\\{PARTOK}=\\{STRUCT}(%
\\{STRING}\|v,\39\\{INT}\\{ty})$;\par
\fi

\M3. Here is the !NIL! declaration.

\Y\P$\4\X1:\X\mathrel{+}\S$\6
$\\{REF}\\{PARTOK}\\{no}\\{tok}=\\{NIL}$; $\\{Lexical}\\{scanning}.\\{The}%
\\{routine}!\\{get}\\{par}\\{tok}!\\{scans}\\{the}\\{stream}$ \&{of} $%
\\{characters}\\{containing}\\{the}\\{parameters}\W\\{yields}\\{the}\\{next}%
\\{token}$ \&{of} $\\{mode}!\\{REF}\\{PARTOK}!,\39\V!\\{no}\\{par}!$  \6
\&{if} $\\{there}\\{are}\\{no}\\{more}\\{parameters}\V\|a\\{lexical}\\{error}%
\\{has}\\{occurred}.$ \6
\&{end} \6
\4$\\{tok}=0\\{open}\\{tok}=1\W\\{also}\\{tok}=2\\{close}\\{tok}=3\\{string}%
\\{tok}=4\\{numb}\\{tok}=5\\{prog}\\{tok}=6\X1:\X=\\{PROC}\\{get}\\{par}%
\\{tok}=(\\{REF}\\{FILE}\|f)\\{REF}\\{PARTOK}$: \37$\\{BEGIN}\\{HEAP}\\{PARTOK}%
\\{hp}\K($$\.{""},\39$$\6
\&{end}\\{tok}$ ) ;\5
\\{CHAR}\\{ch};\5
\\{INT}\|p; \\{on}\\{logical} \&{file}  \6
\&{end} $(\|f,\39(\\{REF}\\{FILE}\\{ff})\\{BOOL}:(\\{GOTO}\\{eof};\,\35%
\\{SKIP}))$;\5
$\X4:Skip blanks and tab characters\X\\{CASE}\\{IF}\\{char}\in\\{string}(%
\\{ch},\39\|p,\39\.{""}\.{"(,)"})\\{THEN}\|p\\{ELSE}0\\{FI}\\{IN}\#\.{"\#\ \ \
@<Get\ a\ string\ denotation@>,}\#(\#\\{hp}\K(\\{ch},\39\\{open}\\{tok}),\39\#,%
\39\#\\{hp}\K(\\{ch},\39\W\\{also}\\{tok}),\39\#)\#\\{hp}\K(\\{ch},\39\\{close}%
\\{tok})\\{OUT}\\{IF}\\{ch}\G\.{"a"}&\\{ch}\L\.{"z"}\\{OR}\\{ch}\G\.{"A"}&%
\\{ch}\L\.{"Z"}\\{OR}\\{char}\in\\{string}(\\{ch},\39\\{LOC}\\{INT},\39\.{"/.%
\~"})\\{THEN}\X6:Process a program identification\X\\{ELIF}\\{ch}\G\.{"0"}&%
\\{ch}\L\.{"9"}\\{OR}\\{ch}=\.{"-"}\\{OR}\\{ch}=\.{"+"}\\{THEN}\X7:Process a
number\X\\{ELIF}\\{ch}=\\{nul}\\{ch}\\{THEN}\\{hp}\\{ELSE}\\{backspace}(\|f)$;\6
\4\\{no}\\{tok}\\{FI}\\{ESAC}\\{EXIT}\\{eof}: \37\\{hp}\\{END};\5
$\#\\{get}\\{par}\\{tok}\#$\par
\fi

\M4. Blanks and tab characters are skipped.

\Y\P$\4\X4:Skip blanks and tab characters\X\S$\6
$\\{WHILE}\\{get}(\|f,\39\\{ch})$;\5
$\\{ch}=\\{blank}\\{OR}\\{ch}=\\{tab}\\{ch}\\{DO}\\{SKIP}\\{OD}$;\par
\U3.\fi

\M5. String denotations can contain embedded doubled string quotes. A
single quote should be inserted if so.

quote = """"

\Y\P$\4\X1:\X\mathrel{+}\S$\6
$\\{BEGIN}\\{make}\\{term}(\|f,\39\\{quote})$;\5
\\{WHILE}\\{STRING}\|s;\5
$\\{get}(\|f,\39(\|s,\39\\{LOC}\\{CHAR},\39\\{ch}))$;\5
$\|v\\{OF}\\{hp}+\K\|s$;\5
$\\{ch}=\\{quote}\\{DO}\|v\\{OF}\\{hp}+\K\\{quote}\\{OD}$;\5
$\\{backspace}(\|f)$;\5
$\\{ty}\\{OF}\\{hp}\K\\{string}\\{tok}$;\5
\\{hp}\\{END}\par
\fi

\M6. A program identification will usually look like an identifier, but
it could be the path to a program. We need to strip whitespace
(blanks and tab characters) from it. In fact any sequence of
characters starting with a lower case letter or a !"/"! or a !"."!
but excluding one of the terminators is acceptable.

\Y\P$\4\X6:Process a program identification\X\S$\6
$\\{WHILE}\\{NOT}\\{char}\in\\{string}(\\{ch},\39\\{LOC}\\{INT},\39\.{"(,)"}+%
\\{nul}\\{ch})\\{DO}\\{IF}\\{ch}/=\\{blank}&\\{ch}/=\\{tab}\\{ch}\\{THEN}\|v%
\\{OF}\\{hp}+\K\\{ch}\\{FI}$;\5
$\\{get}(\|f,\39\\{ch})\\{OD}$;\5
$\\{ty}\\{OF}\\{hp}\K\\{prog}\\{tok}$;\5
$\\{backspace}(\|f)$;\5
\\{hp}\par
\U3.\fi

\M7. We do not try to distinguish between integers and reals, but simply
try to read a real after backspacing once.

\Y\P$\4\X7:Process a number\X\S$\6
$\\{WHILE}\\{char}\in\\{string}(\\{ch},\39\\{LOC}\\{INT},\39\\{tab}\\{ch}+\.{"\
0123456789-+.e"})\\{DO}\\{IF}\\{ch}/=\\{blank}&\\{ch}/=\\{tab}\\{ch}\J\\{THEN}%
\|v\\{OF}\\{hp}+\K\\{ch}\J\\{FI}$;\5
$\\{get}(\|f,\39\\{ch})\\{OD}$;\5
$\\{ty}\\{OF}\\{hp}\K\\{numb}\\{tok}$;\5
$\\{backspace}(\|f)$; \\{hp}\\{Syntactic}\\{analysis}.\\{Parameters}\\{may}%
\\{consist} \&{of} \\{one} \&{of} \6
\4\\{the}\\{following}: $-\$ \6
\&{begin} \37\C{enumerate}\6
$\\\{item}\|A\\{string}\\{denotation}.\\\{item}\|A\\{number}.\\\{item}\|A%
\\{syntactic}\\{token}.\\\{item}\|A$ \6
\4\&{program}\1\  \37$\\{identification}\\{followed}\\{by}\|a\\{parameter}%
\\{pack}.\$ \6
\&{end} \C{enumerate}\6
\|A \6
\4\&{program} \1\  \\{identification}\\{may}\\{start} \6
\&{with} $\|a\\{lower}-$ \6
\&{case} $\\{letter}\V\|a\\{stroke}\V\|a\\{dot}\W\\{may}\\{be}\\{followed}%
\\{by}\\{any}\\{number}$ \1\&{of} $\\{other}\\{characters}\\{other}\\{than}%
\\{an}\\{open}\\{parenthesis}\V\|a\\{comma}\V\|a\\{closing}\\{parenthesis}.\|A%
\\{parameter}\\{pack}\\{consists}$ \&{of} \6
\4$\\{one}\V\\{more}\\{parameters}\\{separated}\\{by}\\{commas}\W\\{surrounded}%
\\{by}\\{parentheses}.\\{For}\\{example}$: \37$-\$\1\6
\&{begin} \37\C{center}\6
$\\\{texttt}$\C{grep("params",ls(-1,"*.c"))}\6
$\$\6
\&{end}\C{center}\2\6
\\{which}\\{contains}\\{the} \6
\4\&{program} \1\  $\\{identification}\\\{verb}+\\{grep}+\\{followed}\\{by}\|a%
\\{parameter}\\{pack}$ \6
\&{with} $\\{two}\\{parameters}.\\{The}\\{first}\\{is}\|a\\{string}%
\\{denotation}.\\{The}\\{second}\\{consists}$ \1\&{of} \\{another} \6
\4\&{program} \1\  $\\{identification}\\\{verb}+\\{ls}+\\{followed}\\{by}\|a%
\\{parameter}\\{pack}$ \6
\&{with} $\\{two}\\{parameters}:\\{an}\\{integer}\W\|a\\{string}\\{denotation}.%
\\{Here}\\{are}\\{some}\\{modes}\\{which}\\{describe}\\{this}\\{structure}.\X1:%
\X=\\{MODE}\\{PARAM},\39\\{VAL}=\\{UNION}(\\{STRING},\39\\{INT},\39\\{REAL},\39%
\\{VOID},\39\\{REF}\\{PARAM}),\39\\{PARAM}=\\{STRUCT}(\\{STRING}\\{prog},\39%
\\{REF}[\,]\\{VAL}\\{val})$;\par
\U3.\fi

\M8. The high-level routine for getting parameters is !get param! which
returns a value of mode !VAL!. This will be !no param! if there are
no parameters.

\Y\P$\4\X1:\X\mathrel{+}\S$\6
$\\{REF}\\{PARAM}\\{no}\\{param}=\\{NIL}$;\par
\fi

\M9. The previous !VAL! is held globally.

\Y\P$\4\X1:\X\mathrel{+}\S$\6
$\\{VAL}\\{prev}\\{val}\K\\{no}\\{param}$;\par
\fi

\M10. Normally, all parameters will be preceded by the identification of
the program or routine.

\Y\P$\4\X1:\X\mathrel{+}\S$\6
\4$\\{PROC}\\{get}\\{param}=(\\{REF}\\{FILE}\|f)\\{VAL}$: \37$\\{IF}\\{REF}%
\\{PARTOK}\\{rp}=\\{get}\\{par}\\{tok}(\|f)$;\5
$\\{rp}\\{IS}\\{no}\\{tok}\\{THEN}\\{no}\\{param}\\{ELSE}\\{CASE}\\{ty}\\{OF}%
\\{rp}\\{IN}\#(\#\\{get}\\{param}(\|f),\39\#,\39\#(\\{prev}\\{val}(\\{REF}%
\\{PARAM}):\\{EMPTY}\\{get}\\{param}(\|f)),\39\#)\#\\{EMPTY},\39\#\\{str}\#%
\\{prev}\\{val}\K\|v\\{OF}\\{rp},\39\#\\{nmb}\#\X13:Get a number\X,\39\#\\{prg}%
\#(\X11:Get a program identification plus parameters\X)\\{OUT}\#$ \6
\&{end} $\#\\{no}\\{param}\\{ESAC}\\{FI}$;\5
$\#\\{get}\\{param}\#$\par
\fi

\M11. A program identification can be followed by arguments.

\Y\P$\4\X11:Get a program identification plus parameters\X\S$\6
$\\{HEAP}\\{PARAM}\\{hp}\K(\|v\\{OF}\\{rp},\39\\{HEAP}[1:0]\\{VAL})$;\5
$\\{prev}\\{val}\K\\{hp}$;\5
$\\{WHILE}\\{VAL}\|v\K\\{get}\\{param}(\|f)$;\6
\4$\\{CASE}\|v\\{IN}(\\{UNION}(\\{INT},\39\\{REAL},\39\\{STRING})\|u)$: \37$(%
\\{hp}+\K\\{VAL}(\|u);\,\35\\{TRUE}),\39(\\{VOID})$: \37$\\{FALSE},\39(\\{REF}%
\\{PARAM}\|p)$: \37$(\|p\\{IS}\\{no}\\{param}\\{FALSE}\\{hp}+\K\\{VAL}(\|p);\,%
\35\\{TRUE})\\{ESAC}\\{DO}\\{SKIP}\\{OD}$;\5
\\{hp}\par
\U10.\fi

\M12. We need to define the operator !+:=! for modes !REF PARAM! and
!VAL!.

\Y\P$\4\X1:\X\mathrel{+}\S$\6
\4$\\{OP}+\K=(\\{REF}\\{PARAM}\\{rp},\39\\{VAL}\|v)\\{VOID}$: ( $\\{HEAP}[1+%
\\{UPB}\\{val}\\{OF}\\{rp}]\\{VAL}\\{hv};\,\35\\{IF}\\{UPB}\\{hv}=1\\{THEN}%
\\{hv}[1]\K\|v\\{ELSE}\\{hv}$ [ : $\\{UPB}\\{hv}-1$ ] $\K\\{val}\\{OF}\\{rp}$;\5
$\\{hv}[\\{UPB}\\{hv}]\K\|v\\{FI}$;\5
$\\{val}\\{OF}\\{rp}\K\\{hv}$ ) ;\par
\fi

\M13. For the sake of portability, we shall use the !stand back! file to
convert a string to a binary integer or real.

\Y\P$\4\X13:Get a number\X\S$\6
$\\{IF}\\{STRING}\|s=\|v\\{OF}\\{rp}$;\5
$\\{char}\in\\{string}(\.{"."},\39\\{LOC}\\{INT},\39\|s)\\{OR}\\{char}\in%
\\{string}(\.{"e"},\39\\{LOC}\\{INT},\39\|s)\\{THEN}\#\\{Process}\|a!\\{REAL}!%
\#\\{REAL}\|r$; \X14:Create a work file and output the string\X\\{on}%
\\{logical} \&{file}  \6
\&{end} $(\\{stand}\\{back},\39(\\{REF}\\{FILE}\|f)\\{BOOL}:(\|r\K0.0;\,\35%
\\{GOTO}\\{eofr};\,\35\\{SKIP}))$;\5
$\\{get}(\\{stand}\\{back},\39\|r)$;\6
\4\\{eofr}: \37$\\{scratch}(\\{stand}\\{back})$;\5
$\\{prev}\\{val}\K\|r\\{ELIF}\\{BOOL}\\{dig}\K\\{FALSE}$;\5
$\\{FOR}\|i\\{FROM}\\{LWB}\|s\\{TO}\\{UPB}\|s\\{WHILE}\\{NOT}(\\{dig}\K\|s[\|i]%
\G\.{"0"}&\|s[\|i]\L\.{"9"})\\{DO}\\{SKIP}\\{OD}$;\5
$\\{NOT}\\{dig}\\{THEN}\#\\{No}\\{digits}\in\\{the}\\{string},\39\\{so}\R\|a%
\\{number}\#\\{prev}\\{val}\K\|s\\{ELSE}\#\\{Process}\\{an}!\\{INT}!\#\\{INT}%
\|n$; \X14:Create a work file and output the string\X\\{on}\\{logical} \&{file}
 \6
\&{end} $(\\{stand}\\{back},\39(\\{REF}\\{FILE}\|f)\\{BOOL}:(\|n\K0;\,\35%
\\{GOTO}\\{eofi};\,\35\\{SKIP}))$;\5
$\\{get}(\\{stand}\\{back},\39\|n)$;\6
\4\\{eofi}: \37$\\{scratch}(\\{stand}\\{back})$;\5
$\\{prev}\\{val}\K\|n\\{FI}$\par
\U10.\fi

\M14. We create a work file using !stand back! and !stand back channel!.

\Y\P$\4\X14:Create a work file and output the string\X\S$\6
$\\{create}(\\{stand}\\{back},\39\\{stand}\\{back}\\{channel})$;\5
$\\{put}(\\{stand}\\{back},\39(\|s,\39\.{"A"}))$;\5
$\\{reset}(\\{stand}\\{back})$;\par
\Us13\ET13.\fi

\M15. Here are the public tags.

\Y\P$\4\X15:Public tags\X\S$\6
$+\K(\\{REF}\\{PARAM},\39\\{VAL}),\39\\{PARAM},\39\\{VAL},\39\\{get}\\{param},%
\39\\{no}\\{param}\\{System}\\{dependencies}.\\{Check}\\{these}$  \6
\&{if} $\\{you}\\{are}\\{porting}\\{this}$ \6
\4\&{program}\1\  \37$\mathrel{\&{to}}\\{another}\\{platform}.$\par

\U2.\fi


\inx
\:\\{also}, 3.
\:\\{an}, 7, 13.
\:\\{analysis}, 7.
\:\\{another}, 7, 15.
\:\\{any}, 7.
\:\\{are}, 3, 7, 15.
\:\\{back}, 13, 14.
\:\\{backspace}, 3, 5, 6, 7.
\:\\{basic}, 2.
\:\\{be}, 7.
\:\\{BEGIN}, 3, 5.
\:\\{blank}, 4, 6, 7.
\:\\{BOOL}, 3, 13.
\:\\{by}, 7.
\:\\{CASE}, 3, 10, 11.
\:\\{ch}, \[3], 4, 5, 6, 7.
\:\\{channel}, 14.
\:\\{CHAR}, 3, 5.
\:\\{char}, 3, 6, 7, 13.
\:\\{characters}, 2, 3, 7.
\:\\{Check}, 15.
\:\\{close}, 3.
\:\\{closing}, 7.
\:\\{comma}, 7.
\:\\{commas}, 7.
\:\\{consist}, 7.
\:\\{consists}, 7.
\:\\{containing}, 3.
\:\\{contains}, 2, 7.
\:\\{CONTEXT}, 1.
\:\\{create}, 14.
\:\\{DECS}, 1.
\:\\{denotation}, 7.
\:\\{dependencies}, 15.
\:\\{describe}, 7.
\:\\{determines}, 2.
\:\\{dig}, \[13].
\:\\{digits}, 13.
\:\\{DO}, 4, 5, 6, 7, 11, 13.
\:\\{dot}, 7.
\:\\{ELIF}, 3, 13.
\:\\{ELSE}, 3, 10, 12, 13.
\:\\{EMPTY}, 10.
\:\\{END}, 3, 5.
\:\\{eof}, 3.
\:\\{eofi}, 13.
\:\\{eofr}, 13.
\:\\{error}, 3.
\:\\{ESAC}, 3, 10, 11.
\:\\{example}, 7.
\:\\{EXIT}, 3.
\:\|{f}, \[3], \[13].
\:\\{FALSE}, 11, 13.
\:\\{ff}, \[3].
\:\\{FI}, 3, 6, 7, 10, 12, 13.
\:\\{field}, 2.
\:\\{FILE}, 3, 10, 13.
\:\\{FINISH}, 2.
\:\\{first}, 7.
\:\\{followed}, 7.
\:\\{following}, 7.
\:\\{FOR}, 13.
\:\\{For}, 7.
\:\\{FROM}, 13.
\:\\{get}, 3, 4, 5, 6, 7, 10, 11, 13, 15.
\:\\{GOTO}, 3, 13.
\:\\{grep}, 7.
\:\\{has}, 3.
\:\\{HEAP}, 3, 11, 12.
\:\\{Here}, 7.
\:\\{hp}, \[3], 5, 6, 7, \[11].
\:\\{hv}, \[12].
\:\\{identification}, \[7].
\:\\{IF}, 3, 6, 7, 10, 12, 13.
\:\\{IN}, 3, 10, 11.
\:\\{INT}, 2, 3, 6, 7, 11, 13.
\:\\{integer}, 7.
\:\\{is}, 2, 7.
\:\\{IS}, 10, 11.
\:\\{item}, 7.
\:\\{KEEP}, 2.
\:\\{letter}, 7.
\:\\{lexical}, 3.
\:\\{Lexical}, 3.
\:\\{LOC}, 3, 5, 6, 7, 13.
\:\\{logical}, 3, 13.
\:\\{lower}, 7.
\:\\{ls}, 7.
\:\\{LWB}, 13.
\:\\{make}, 5.
\:\\{may}, 7.
\:\\{mode}, 2, 3.
\:\\{MODE}, 2, 7.
\:\\{modes}, 7.
\:\\{more}, 3, 7.
\:\|{n}, \[13].
\:\\{next}, 3.
\:\\{NIL}, 3, 8.
\:\\{nmb}, 10.
\:\\{no}, \[3], \[8], \[9], 10, 11, 15.
\:\\{No}, 13.
\:\\{NOT}, 6, 13.
\:\\{nul}, 3, 6.
\:\\{numb}, 3, 7.
\:\\{number}, 7, 13.
\:\\{occurred}, 3.
\:\\{OD}, 4, 5, 6, 7, 11, 13.
\:\\{OF}, 5, 6, 7, 10, 11, 12, 13.
\:\\{on}, 3, 13.
\:\\{one}, 7.
\:\\{OP}, 12.
\:\\{open}, 3, 7.
\:\\{OR}, 3, 4, 13.
\:\\{other}, 7.
\:\\{OUT}, 3, 10.
\:\|{p}, \[3], \[11].
\:\\{pack}, 7.
\:\\{par}, 3, 10.
\:\\{param}, 8, 9, 10, 11, 15.
\:\\{PARAM}, \[7], 8, 10, 11, 12, 15.
\:\\{parameter}, 7.
\:\\{parameters}, 3, 7.
\:\\{Parameters}, 7.
\:\\{params}, 1.
\:\\{parentheses}, 7.
\:\\{parenthesis}, 7.
\:\\{PARTOK}, 2, 3, 10.
\:\\{platform}, 15.
\:\\{porting}, 15.
\:\\{prev}, \[9], 10, 11, 13.
\:\\{prg}, 10.
\:\\{PROC}, 3, 10.
\:\\{Process}, 13.
\:\\{prog}, 3, 6, 7.
\:\\{put}, 14.
\:\\{quote}, 5.
\:\|{r}, \[13].
\:\\{REAL}, 7, 11, 13.
\:\\{REF}, 3, 7, 8, 10, 11, 12, 13, 15.
\:\\{reset}, 14.
\:\\{routine}, 3.
\:\\{rp}, \[10], 11, \[12], 13.
\:\|{s}, \[13].
\:\\{scanning}, 3.
\:\\{scans}, 3.
\:\\{scratch}, 13.
\:\\{second}, 7.
\:\\{separated}, 7.
\:\\{SKIP}, 3, 4, 11, 13.
\:\\{so}, 13.
\:\\{some}, 7.
\:\\{stand}, 13, 14.
\:\\{standard}, 1.
\:\\{start}, 7.
\:\\{str}, 10.
\:\\{stream}, 3.
\:\\{STRING}, 2, 5, 7, 11, 13.
\:\\{string}, 3, 5, 6, 7, 13.
\:\\{stroke}, 7.
\:\\{STRUCT}, 2, 7.
\:\\{structure}, 2, 7.
\:\\{surrounded}, 7.
\:\\{Syntactic}, 7.
\:\\{syntactic}, 7.
\:\\{System}, 15.
\:{system dependencies}, 1, 2.
\:\\{tab}, 4, 6, 7.
\:\\{term}, 5.
\:\\{texttt}, 7.
\:\\{than}, 7.
\:\\{the}, 2, 3, 7, 13.
\:\\{The}, 2, 3, 7.
\:\\{THEN}, 3, 6, 7, 10, 12, 13.
\:\\{there}, 3.
\:\\{these}, 15.
\:\\{this}, 7, 15.
\:\&{to}, \[15].
\:\\{TO}, 13.
\:\\{tok}, 3, 5, 6, 7, 10.
\:\\{token}, 2, 3, 7.
\:\\{Tokens}, 2.
\:\\{tokens}, 2.
\:\\{TRUE}, 11.
\:\\{two}, 7.
\:\\{ty}, 2, 5, 6, 7, 10.
\:\|{u}, \[11].
\:\\{UNION}, 7, 11.
\:\\{UPB}, 12, 13.
\:\\{USE}, 1.
\:\|{v}, \[11], \[12].
\:\\{val}, 7, 9, 10, 11, 12, 13.
\:\\{VAL}, \[7], 9, 10, 11, 12, 15.
\:\\{verb}, 7.
\:\\{VOID}, 1, 7, 11, 12.
\:\\{which}, 2, 7.
\:\\{WHILE}, 4, 5, 6, 7, 11, 13.
\:\\{yields}, 3.
\:\\{you}, 15.
\fin
\:\X14:Create a work file and output the string\X
\Us13\ET13.
\:\X13:Get a number\X
\U10.
\:\X11:Get a program identification plus parameters\X
\U10.
\:\X7:Process a number\X
\U3.
\:\X6:Process a program identification\X
\U3.
\:\X15:Public tags\X
\U2.
\:\X4:Skip blanks and tab characters\X
\U3.
\con
